Designing Types for R, Empirically

### 2020-10-15 paper/main.tex
wip

Given the pecularities of the language there are a number of design
choices that need to be reviewed and evaluated. It is uncontroversial
to include types that cover the six kinds of primitive vectors,
furthermore environments and the distinguished null types are commonly
used and must be included. For simplicity, we omit some of the data
types that closer to the implementation of the
language. Fig.~\ref{types} presents our type language.

@@

\paragraph{Scalars} While R does not have scalar data types, there are
cases where functions except scalar values, for example a conditional
takes a single logical (and will complain if more values are passed).
We considered tracking dimensions of data structures, but decided
against it. Instead, the type language differentiates between vectors
of lenght 1 and vectors of any dimension.  The primitive types can be
either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
happen to be of length 1, and thus a scalar is also a vector.  Vectors
are monomorphic, a vector of doubles contains only doubles.

@@

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$. It is noteworthy that the elements
of a union are not necessarily disjoint. For instance a \NAVEC(\int)

\paragraph{Nullables} 
The \nul type is inhabited by a singleton \code{NULL} value often used
as a sentinel or a stand in for arguments with no value.  To capture
this behavior, we introduce a nullable type \code{?}\T.  The
functional difference between the \NA and \code{NULL} values is that
the latter cannot be stored in vectors.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list with
\code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). The list type is parameterize: \LIST\T.

\paragraph{Classes} 
R has more than one notion of type. As we mentioned, values can be
attributed, and a very important attribute is the class of a value.  A
class is a list of names that are used to mimic object-oriented
programming. The type system includes class types written
\CLASS{\ID_1,\dots,\ID_n}.

\paragraph{Environments}
Environments are lists with reference semantics: mutating a value in an
environment is performed in-place.  They are used to store variables and to
escape from the copy-on-write semantics of other data types.

\paragraph{Functions.}
Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
$\A_i$ argument is either a type \T or \VARGS, a variable length
argument list. Moreover, a single function's signature can be the
disjunction of a number of individual signatures.

\section{Extracting and Checking Signatures}

@@

\begin{figure}[t!]
  \vspace{2mm}
  \includegraphics[width=\textwidth]{plots/analysis-pipeline-new.pdf}
  \caption{The Analysis Pipeline. (1) the rather long time is because
    we had to extract metadata from all CRAN packages before we could
    compose our \PACKAGES package corpus. Here CRAN refers to the
    Comprehensive R Archive Network, R's repository of packages and
    libraries.}
  \label{fig:pipeline}
\end{figure}

@@

\subsection{Inferring type signatures from raw types}
\enlargethispage{4mm}
The output of the \typetracer tool consists of a set of tuples of raw types,
each representing a function call in some program's execution. The inference
step consolidates the different tuples corresponding to a particular function
definition across multiple programs and distills them into a single type
signature.  The shape of the inferred function signatures is:

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function
types. Furthermore, we apply some transformation on the types to keep
the size of types in check.  Figure~\ref{fig:siplification-rules}
overviews the main simplification rules as well as the simple notion
of subtyping that we have adopted here.  Our subtyping rules are both
born out of necessity (e.g., scalars are subtypes of vectors since
a. that is the case in R and b. we want to allow types like \dbl |
\VEC\dbl to simplify), as well as from prevalent data type conversion
with no ambiguity (e.g., \int to \dbl).

\begin{figure}[h!tb]
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\nul \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
  \\
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\caption{Simplification rules and subtyping}
\label{fig:siplification-rules}
\end{figure}

@@

% \newpage

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

Given the pecularities of the language there are a number of design
choices that need to be reviewed and evaluated. It is not
controversial to include types that cover the six kinds of primitive
vectors, furthermore environments and the distinguished null types are
commonly used and must be included.  Environments are lists with
reference semantics: mutating a value in an environment is performed
in-place.  They are used to store variables and to escape from the
copy-on-write semantics of other data types.  For simplicity, we omit
some of the data types that closer to the implementation of the
language. Fig.~\ref{types} presents our type language.

@@

\paragraph{Scalar} While R does not have scalar data types, there are
cases where functions except scalar values, for example a conditional
takes a single logical (and will complain if more values are passed).
We considered tracking dimensions of data structures, but decided
against it. Instead, the type language differentiates between vectors
of lenght 1 and vectors of any dimension.  The primitive types can be
either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
happen to be of length 1, and thus a scalar is also a vector.  Vectors
are monomorphic, a vector of doubles contains only doubles.

@@

\paragraph{Nullable} 
The \nul type is inhabited by a singleton \code{NULL} value often used
as a sentinel. Unlike in some other lanugage, \k{NULL} is not the
default value of uninitialized variables. R has different notion for
that (which we do not cover here).  To capture common uses, of
\k{NULL}, the type language has a nullable type, written \code{?}\T.
Values of this type can be either values of type \T or \k{NULL}.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list
with \code{as.list}, and lists to vectors with \code{unlist}
(coercions may ensue). The type language allows to specify that a
value is a list containing element of some type \T, written \LIST\T. R
does not have built-in type tests for this purpose, to establish the
type of a value requires traversal and checking individual elements.

\paragraph{Class}
R has three objects systems, code-named S3, S4 and R5. All of them
operate by adding a class attribute to values. The most widely used
system, by far, is S3 which supports single dispatch and multiple
classes~\cite{ecoop12}. The challenge from a type system point of view
is that a value such as integer 5, could be attributed with a class.
Code that performs dispatch would use the class attribute while code
that does not would view the value as an integer. The type language
focuses on the attribute and will hide the underlying type of the
value. While that seems to match common usage it does represent a loss
of expressiveness. The type language also focuses on S3, we leave the
other object systems for future work. S3 has no notion of inheritance,
each value has a list of classes. The type language thus allows to
write \CLASS{\ID_1,\dots,\ID_n} to denote values tagged with the class
names $\ID_1$ to $\ID_n$.

\paragraph{Union} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$. The elements of unions are not
disjoint, 1L is both an \int, a \VEC\int and a \NAVEC\int. R does not
provide a built-in type testing mechanism, e.g. for the case of
\NA-free data types, it is necessary to scan vectors to find out if
they have missing values.

\paragraph{Function}
Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
$\A_i$ argument is either a type \T or \VARGS, a variable length
argument list. Moreover, a single function's signature can be the
disjunction of a number of individual signatures.

\subsection{Subtyping}

The choice of type language follows the structure of values.  The
presence of \NA-free data types and scalars are two choices that must
be validated in practice.  Nullable types are just a special case of
unions.

While R does not support the notion of subtyping between values.  The
conversions between primitive types give us a starting point (e.g., an
\int is always accepted where a \dbl is expected). Furthermore, the
types introduced above induce some the rules of Figure~\ref{fig:sub}:
a vector without \NA is a subtype of a vector, a value of type \T is a
subtype of a nullable \T, a list is subtype of another list if their
elements are subtypes, and a scalar is a subtype of a vector of the
same primitive type.

\begin{figure}[h!tb]\begin{tabular}{rclll}
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\caption{Subtyping}\label{fig:sub}
\end{figure}

\subsection{Synthesizing signatures}

The output of the \typetracer tool consists of a set of tuples of raw
types, each representing a function call in some program's
execution. The inference step consolidates the different tuples
corresponding to a particular function definition across multiple
programs and distills them into a single type signature.  The shape of
the inferred function signatures is:

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function
types. Furthermore, we apply some transformation on the types to keep
the size of types in check.  Figure~\ref{fig:siplification-rules}
overviews the main simplification rules as well as the simple notion
of subtyping that we have adopted here.

\begin{figure}[h!tb]\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\nul \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
\end{tabular}
\caption{Simplification rules}\label{fig:siplification-rules}
\end{figure}

\section{Type Extraction and Checking Pipeline}

@@

\begin{figure}[h!]  \vspace{2mm}
  \includegraphics[width=\textwidth]{plots/analysis-pipeline-new.pdf}
  \caption{The Analysis Pipeline; $^1$ metadata has to be extracted
    for \emph{all} CRAN packages.}\label{fig:pipeline}
\end{figure}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
